%!TEX root = depicto-top.tex
\chapter*{Acknowledgements}

First and foremost, I would like to thank my supervisor, \emph{Vincent
Vandeghinste}, for giving me the freedom to determine the scope of my research
while also being there when I needed help dialing it in. Needless to say, I am
very grateful to him for sacrificing so many of his Friday afternoons these
past few months. Our meetings, which, I swear, I really did go into with all
the intentions of brevity, were a consistent source of encouragement and did
much in the way of answering those questions which I could not articulate for
myself. Also present at most of these meetings was \emph{Leen Sevens}, whom I
would like to thank from the bottom of my heart for her insight and, perhaps
most of all, enthusiasm. In return, I have taken care to give as adequate a
representation of the \emph{Picto2Text} system as my own ignorance will permit.
However, should I have misrepresented any part of the system, I apologise in
advance. I would also like to thank the other partners at the Centre for
Computational Linguistics: \emph{Ineke Schuurman}, who, throwing an observation
over her computer monitor from time to time, sometimes pushed the conversation
in exciting new directions, and always made me feel welcome; and \emph{Prof.
Frank Van Eynde}, for helping me find my way around all the subtly different
flavors of \hpsg, and for his lectures on \hpsg\ (attended last year), without
which the work presented here would most likely have taken a very different,
and most likely bumpy, course. On a similar note, I would also like to thank
\emph{Prof. Piet Mertens}, whose master's course on parsing (also attended last
year) provided me with enough frame of reference to get to grips with the
\delphin\ formalism quickly and independently.

Next, I would like to express my gratitude to all contributors to the \delphin\
project, with special thanks to \emph{Woodley Packard} for developing (and
open-sourcing) the \ace\ processor, as well as the maintainers of the \delphin\
wiki, where, for the past months, I have practically lived. Thank you,
maintainers, for creating such a sense of community.

Last but certainly not least, I would like to thank those dear to me, starting
with my parents, to whom I am beyond grateful, not only for putting up with me
and supporting my ambitions, but for providing me with the opportunity to spend
so many months focusing solely on my research. Further, I would like to thank
my brother and sister, \emph{Reinert} and \emph{Evelien}, as well as \emph{Eva
Vanermen}, \emph{Sue Goossens}, \emph{Xavier G.A.},  \emph{Fred Giannetti}, and
\emph{Charlotte} `discosis' \emph{Tesselaar}, for their truly one-of-a-kind
company and for reminding me from time to time not to take myself too
seriously. Finally, I would like to thank my girlfriend, \emph{Elisa Maes}.
More than anyone else these past few months, she has kept me grounded when I
wanted to float away, fascinating and frustrating me every day in a way that I
cherish, but cannot explain, being \emph{her} and no-one else, for which I
thank her profoundly.
